---
title: "Brushing up on the linear model"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1
The general linear model: Y=Xβ+ϵ:
Do a linear regression, expressing fuel usage as a function of weight using the function lm
```{r}
data<-mtcars
model <- lm(data[,1]~data[,6])
```

1. Extract β^, Y, Y^, X and ϵ from model (hint: have a look at the function model.matrix)

```{r}

```

i. create a plot that illustrates Y and Y^ (if you are feeling ambitious, also include ϵ (hint: you can use the function arrows))

```{r}
plot(data[,1],data[,6])
```

2. Estimate β for a quadratic model (y=β2x2+β1x+β0) using ordinary least squares without using lm; β^=(XTX)−1XTY (hint: add a third column to X from step 1)

3. Compare your acquired β^ with the output of the corresponding quadratic model created using lm (hint: use the function I, see details under help and the sub-section formula operators here: https://www.datacamp.com/community/tutorials/r-formula-tutorial)
create a plot that illustrates Y and Y^ (if you are feeling ambitious, also include ϵ (hint: you can use the function arrows))
